---
title: "GEOG3023_Tally(O)_ANOVA_AND_CHISQ_TESTs"
author: "Kristin Robinson"
date: "Dec.2 , 2016"
output: html_document
---

---

This document shows how to access storm data from Google sheets, "clean"" the data and merge population data from another census data. 
The data stored in the Google sheet has all the data from 2000-2015 from the NOAA "storm details" and "fatalities" csv files. 
You must install and include the gsheet library to access the Google sheet.
There are some sample charts and graphs in the last R chunk. These are just for example and can be removed.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
```

```{r}
##LOAD LIBRARIES AND LOAD DATA FROM GOOGLE DOCS

# Libraries
library(mosaic)
library(gsheet)
library(ggplot2)
library(plyr)
library(pscl)

# Data
 x <- getURL("https://raw.githubusercontent.com/RobinsonK/GEOG3031/master/storm2.csv?token=ARoUsTaimJLs9T1Bo4LPTMTgdoP6EUMGks5YSyWAwA%3D%3D")
x2<-read.csv(text=x)
storm<-aggregate(cbind(FATALITIES_COUNT) ~ EVENT_ID + STATE+MONTH_NAME + EVENT_TYPE +FATALITY_LOCATION+POP_PERFAT, data = x2, sum)
```



```{r}

mod<-glm(formula = FATALITIES_COUNT ~ EVENT_TYPE +STATE +MONTH_NAME+EVENT_TYPE+FATALITY_LOCATION+POP+PERFAT, family = poisson(link=log), data = storm)

summary(mod)
summary(mod)$coefficients
pR2(mod)
anova(mod, test="Chisq")

```

