---
title: "CleaningDataandMergingwithCensusPop"
author: "Kristin Robinson"
date: "November 7, 2016"
output: html_document
---

---

This document shows how to access storm data from Google sheets, "clean"" the data and merge population data from another census data. 
The data stored in the Google sheet has all the data from 2000-2015 from the NOAA "storm details" and "fatalities" csv files. 
You must install and include the gsheet library to access the Google sheet.
There are some sample charts and graphs in the last R chunk. These are just for example and can be removed.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE)
```

```{r}
##LOAD LIBRARIES AND LOAD DATA FROM GOOGLE DOCS

# Libraries
library(mosaic)
library(gsheet)
library(ggplot2)
library(plyr)

# Data
url<-'https://docs.google.com/spreadsheets/d/1kjHLs0FaapCnuoS54jHv2RM7UnzIoNl7kJL5uBQyz48/edit?usp=sharing'
storm <- gsheet2tbl(url)
url2<-'https://docs.google.com/a/colorado.edu/spreadsheets/d/1pFAtBDRLIVlZespT92fWgqWYh_6wwZe-a67C7KGz7uU/edit?usp=sharing'
census<-gsheet2tbl(url2)


```

```{r}
##DATA CLEANUP
# 1. Remove columns you do not need.
# 2. Remove all cases with NA, "Unknown" or "Other"
# 3. Add a column for fatalities count
# 4. Combine overlapping event types (ex. combine extreme heat and heat, hurricane/typhoon and hurricane, etc.) NOTE: These are suggestions and can be easily changed!



# Choose columns to get rid of
drop <- c("X","BEGIN_YEARMONTH","BEGIN_DAY","BEGIN_TIME","END_YEARMONTH","END_DAY","END_TIME","EPISODE_ID","STATE_FIPS","CZ_TYPE","CZ_FIPS","CZ_NAME","WFO","BEGIN_DATE_TIME","CZ_TIMEZONE","END_DATE_TIME","INJURIES_DIRECT","INJURIES_INDIRECT","DEATHS_DIRECT","DEATHS_INDIRECT","DAMAGE_PROPERTY","DAMAGE_CROPS","SOURCE","MAGNITUDE","MAGNITUDE_TYPE","FLOOD_CAUSE","CATEGORY","TOR_F_SCALE","TOR_LENGTH","TOR_WIDTH","TOR_OTHER_WFO","TOR_OTHER_CZ_STATE","TOR_OTHER_CZ_FIPS","TOR_OTHER_CZ_NAME","BEGIN_RANGE","BEGIN_AZIMUTH","BEGIN_LOCATION","END_RANGE","END_AZIMUTH","END_LOCATION","BEGIN_LAT","BEGIN_LON","END_LAT","END_LON","EPISODE_NARRATIVE","EVENT_NARRATIVE","DATA_SOURCE","FAT_TIME","FATALITY_DATE","EVENT_YEARMONTH","FATALITY_TYPE")

# REMOVE UNUSED COLUMNS (VARS)
storm = storm[,!(names(storm) %in% drop)]

# REMOVE CASES W/NA, OTHER and UNKNOWN
storm$FATALITY_LOCATION[storm$FATALITY_LOCATION=="Unknown"] <- NA
storm$FATALITY_LOCATION[storm$FATALITY_LOCATION=="Other"] <- NA
storm$FATALITY_SEX[storm$FATALITY_SEX=="Unknown"] <- NA
storm$FATALITY_SEX[storm$FATALITY_SEX==""] <- NA
storm <- na.omit(storm)


# ADD FATALITIES COUNT COLUMN
storm$FATALITIES_COUNT <- 1  # Use the same value (1) for all rows

# CONDSENSE CATEGORIES
storm$EVENT_TYPE[grep("*Astronomical Low Tide*|*High Surf*|*Rip Current*|*Storm Surge/Tide*|*Sneakerwave*|*Tsunami*", storm$EVENT_TYPE)] <- "Tide/Tsunami"
storm$EVENT_TYPE[grep("*Blizzard*|*Heavy Snow*|*Ice Storm*|*Lake-Effect Snow*|*Winter Weather*|*Winter Storm*", storm$EVENT_TYPE)] <- "Winter Storm"
storm$EVENT_TYPE[grep("*Coastal Flood*|*Flash Flood* |*Flood*", storm$EVENT_TYPE)] <- "Flood"
storm$EVENT_TYPE[grep("*Cold/Wind Chill*|*Extreme Cold/Wind Chill*|*Frost/Freeze*|*Sleet*|*Cold*", storm$EVENT_TYPE)] <- "Cold"
storm$EVENT_TYPE[grep("*Marine High Wind*|*Marine Strong Wind*|*Marine Thunderstorm Wind*|*Waterspout*|*Marine Dense Fog*", storm$EVENT_TYPE)] <- "Marine Weather"
storm$EVENT_TYPE[grep("*Debris Flow*|*Landslide*|*Avalanche*", storm$EVENT_TYPE)] <- "Landslide"
storm$EVENT_TYPE[grep("*Freezing Fog*|*Dense Fog*", storm$EVENT_TYPE)] <- "Fog"
storm$EVENT_TYPE[grep("*Dense Smoke*|*Wildfire*", storm$EVENT_TYPE)] <- "Fire"
storm$EVENT_TYPE[grep("*Dust Devil*|*Dust Storm*", storm$EVENT_TYPE)] <- "Dust Storm"
storm$EVENT_TYPE[grep("*High Wind*|*Strong Wind*|*Thunderstorm Wind*", storm$EVENT_TYPE)] <- "Wind"
storm$EVENT_TYPE[grep("*Excessive Heat*|*Heat*", storm$EVENT_TYPE)] <- "Heat"
storm$EVENT_TYPE[grep("*Hail*", storm$EVENT_TYPE)] <- "Hail"
storm$EVENT_TYPE[grep("*Heavy Rain*", storm$EVENT_TYPE)] <- "Heavy Rain"
storm$EVENT_TYPE[grep("*Lightning*", storm$EVENT_TYPE)] <- "Lightning"
storm$EVENT_TYPE[grep("*Hurricane*|*Hurricane (Typhoon)*|*Tropical Depression*|*Tropical Storm*", storm$EVENT_TYPE)] <- "Hurricane"
storm$EVENT_TYPE[grep("*Tornado*|*Funnel Cloud*", storm$EVENT_TYPE)] <- "Tornado"

#CONDENSE LOCATIONS
storm$FATALITY_LOCATION[grep("*Long Span Roof*", storm$FATALITY_LOCATION)] <- "Roof"
storm$FATALITY_LOCATION[grep("*Permanent Structure*", storm$FATALITY_LOCATION)] <- "Perm. Struct"
storm$FATALITY_LOCATION[grep("*Outside/Open Areas*", storm$FATALITY_LOCATION)] <- "Open Space"
storm$FATALITY_LOCATION[grep("*Heavy Equipment/Construction*", storm$FATALITY_LOCATION)] <- "Heavy Equipment"
storm$FATALITY_LOCATION[grep("*Boat*|*Boating*", storm$FATALITY_LOCATION)] <- "Boat"
storm$FATALITY_LOCATION[grep("*Mobile/Trailer Home*", storm$FATALITY_LOCATION)] <- "Mobile Home"
storm$FATALITY_LOCATION[grep("*Vehicle/Towed Trailer*", storm$FATALITY_LOCATION)] <- "Vehicle"
storm$FATALITY_LOCATION[grep("*Permanent Home*", storm$FATALITY_LOCATION)] <- "Perm. Home"

# Toupper state names in State Pop data
census$STATE<-toupper(census$STATE)
# Add population data to storm dataframe
storm <- join(storm, census, by = "STATE")

storm <-na.omit(storm)
```

```{r}
##SAMPLE GRAPHS AND CHARTS (very basic data)

#1. Histogram: Fatalities per Year
hist(storm$YEAR, main="Number of Fatalities per Year", xlab="Year", ylab="Fatalities")

#2. Barplot and List: Fatalities by event type
ET<-table(storm$EVENT_TYPE)
ET = as.data.frame(ET)
ET = ET[order(ET$Freq, decreasing=TRUE),]
barplot(ET[,2], names.arg = ET[,1], main="Fatalities by Event Type", ylab="Fatalities")
ET

#3. Boxplot and Count: Sex
boxplot(storm$FATALITY_AGE~storm$FATALITY_SEX, main="Age and Sex Fatalities", ylab="Age")
Sex_count<-count(storm$FATALITY_SEX)
Sex_count

#4. Boxplot: Fatality location and age
boxplot(FATALITY_AGE~FATALITY_LOCATION, main="Fatality Location", ylab="Age", data=storm)

```
